Busch, Khakhalin. Intrinsic temporal tuning

### 2019-04-19 reviews and responses.md

> - The rationale for the use of dynamic clamp and the parameters chosen are not clear.

Previously a detailed justification for the use of dynamic clamp was only given in the Discussion; we now added it to the end of Introduction.

> Is the dynamic clamp current injection protocol mimicking the physiological synaptic
input? This should be stated.

We updated the next sentence, by adding a reference to it: "We used conductances of 4 different durations (100, 200, 500, and 1000 ms), and 3 different amplitudes (peak conductances of 0.2, 0.5, and 1.0 nS), matching the range of synaptic currents observed in tectal circuits in vivo (Xu et al. 2011; Khakhalin et al. 2014; Ciarleglio et al. 2015; see Methods)."

@@

> - I think the experiment in this study is pretty straightforward, but the description of the
results is not. As far as I can tell, tadpoles were exposed to different stimuli (flash,
looming, multisensory, etc.). Then, tectal neurons were stimulated using either 1)
dynamic clamp or 2) more physiologically through optic chiasm. This allowed the
investigators to assess firing when they changed the stimulus intensity and stimulus
duration of the dynamic clamp or the synaptic response to presynaptic stimulation.
They then looked how the different types of sensory experience influenced firing
responses. The rest is data analysis that looks at how the duration, amplitude, total
spiking, and synaptic input is changed in different sensory contexts and sensory
stimuli might interact. I think the text can be made much simpler so the basic design
is easier to grasp. I suggest thinking about what the take home point is for each
figure and craft the results section around those messages.

Thank you; we tried to follow this advice in this revised version of the manuscript!

> - There is also an inconsistent use terminology. For example, “spikiness” and
“excitability” are sometimes used in attempt to describe distinct cell characteristics,
but sometimes it seems like they are being used interchangeably. This makes
interpreting the written results difficult to follow.

In this updated version, we consistently use "spikiness" as a name for the variable, for the first time introduced in Figure 2C. For earlier results section, we replaced "spikiness" with "average number of spikes", and in all instances when we talked about the general idea of neuronal excitability, we called it neuronal excitability. As the only exception to this rule, we opted for the word "spikness" in the first paragraph of Introduction, to prime the reader for the use of this word later in the text.

@@

> - “Compared to control, neurons from animals exposed to visual stimuli had a flatter
amplitude tuning curve (Fig 1D,E).” What is the reader supposed to be looking at in
the figure? Is this referring to larger conductances for a fixed duration? I think this is
what the authors’ mean, but more intuitive descriptions that matches the figures or
vice versa is needed to make the whole results section easier to digest.

This refers to smaller (flat, not increasing) spiking output as a function of conductance duration. To make it clear, we improved the sentence:

"Compared to control, neurons from animals exposed to visual stimuli had a flatter **(constant or hill-shaped, rather than increasing)** amplitude tuning curve (Fig 1D,E), **as they** did not increase their spiking as fast in response to larger conductances..."

> - Excessive wordiness makes the results difficult to read and interpret. For example,
on pg. 9, 4 paragraphs (411 words) are used to describe two graphs, each showing
a different linear relationship (Fig. 3E-F). Why not simply say, “A positive relationship
between temporal tuning and synaptic duration implies a co-tuning of preferr
parameters.”

Of these four paragraphs, the last one reports a negative result that is not shown in Figure 3. The first three paragraphs are indeed dedicated to panels Figure 3EF, but we cannot replace them with one sentence, as the statement we are making is not that obvious statistically. We believe that an attempt to gloss it over by pushing the statistical analysis to the methods section would not be fair.

The issue, as we describe in the text, is that while graphs 3E and 3F have very similar axes, and both show a linear dependency, they have opposite slopes. Had the slopes be the same, we would have just reported a correlation, in one sentence. But with the data we have, we need to be more specific, as we observed that the within-group correlation is positive, yet the between-group correlation is negative.

We indeed received some conflicting advice in the past about the level of detail necessary to communicate this message: for example, people who have studied Simpson's paradox in their statistics classes had found our use of this term very helpful, while people who haven't heard it before, found it distracting. For now, we opted to include it, but we are aware that what makes the text more relatable to one group of researchers may make it less transparent for other groups.

Nevertheless, in current version of the paper, we shortened this section from 411 to 294 words.

@@

All our measurements of the number of spikes relied on the dynamic clamp experiments, that used simulated, and not real synaptic inputs. It means that any changes in spiking we observed in the previous section of the results is necessarily coming from the change in intrinsic properties. While a change in the synaptic strength in response to sensory stimulation is possible, and indeed is investigated below in this paper, it would not have affected the number of spikes we reported in figures 1 and 2, as they were entirely recorded in dynamic clamp experiments.

To make this clearer to the reader, we TODO!

> Page 5 line 15: Does the word “preferred” (in quotes) mean that the cells fired more action potentials?
Maybe refrain from personification.

We hope that the word "prefer" is intuitive enough to most readers; it is also way shorter than "responded with more spikes", or "produced more action potentials", which allows for easier writing. We also believe that most readers of this article will not be at risk of truly personifying neurons, so while we take care to never personify animals, we are usually a bit more relaxed when talking about cells, channels, receptors etc.

Still, to make this sentence better, we removed the quotes around "preferred", to make it more obvious that it is just our normal word use (we use this word 6 times in the paper), and not an allegory.

### 2019-04-19 main.tex
Methods: around figure 3

Our first question was whether our stimulation protocols caused changes in intrinsic excitability of tectal neurons. From previous studies, we knew that in tadpoles exposed to four hours of LED flashes, tectal neurons became more excitable \citep{aizenman2003,ciarleglio2015}. However, the stimuli we used in the present study were weaker, and similar to those used in behavioral experiments \citep{khakhalin2014,james2015,truszkowski2017}. We presented a checkerboard pattern that inverted once a second either instantaneously (Figure 1C left, dubbed “Flash”), or with a slow transition over the course of a second (old black squares shrank to white, while new black squares grew from old white squares; Figure 1C right; dubbed “Looming”; see Methods).

@@

After conditioning, we excised the brain, obtained whole cell patch clamp recordings from neurons in the tectum (Figure 1A), and counted spikes produced during dynamic clamp in response to simulated synaptic conductances of different durations and amplitudes (Figure 1B), mimicking synaptic conductances typically observed in tadpole brains (see Methods). We used conductances of 4 different durations (100, 200, 500, and 1000 ms), and 3 different amplitudes (peak conductances of 0.2, 0.5, and 1.0 nS), matching the range of synaptic currents observed in tectal circuits \textit{in vivo} (see Methods). Contrary to our expectations, in tadpoles that were exposed to instantaneous checkerboard inversions (flashes), tectal neurons on average became not more, but less spiky, and generated 0.4$\pm$0.4 spikes, across all types of dynamic clamp inputs, compared to 0.9$\pm$1.0 spikes in control (Figure 1D, left column; F(1,677)=30.4, p=5e-8, n cells=28, 29. Here and below, F-values are reported for a multivariate fixed effects analysis of variance with selected interactions, where cell id is used as a repeated measures factor; see Methods for a detailed description). Neurons exposed to looming transitions also spiked less than control neurons (0.6$\pm$0.4 spikes; F(1,629)=9.6; p=0.002; n=28, 25), but more than those exposed to instantaneous "flashes" (F(1,641)=3.7; p=4e-6; n=29, 25), which was likewise unexpected. In the tadpole tectum, looming stimuli are known to elicit stronger responses compared to instantaneous inversions \citep{khakhalin2014}, yet the suppression yielded by looming stimuli was weaker.

We then mapped the amplitude tuning of neurons (amplitude transfer function, or gain), by looking at how an increase in transmembrane conductance translated into increased spike output. Compared to control, neurons from animals exposed to visual stimuli had a flatter amplitude tuning curve (Fig 1D,E), and did not increase their spiking as fast in response to larger conductances (F(1,677)=15.9, p=8e-5; and F(1,629)=8.3, p=0.004 for flash- and looming transitions respectively). The flattening of response curves seemed slightly more pronounced for flashes than for looming, but this difference was not significant (F(1,641)=3.5; p=0.06).

@@

Across input conductances of different lengths (100, 200, 500, and 1000 ms), neurons exposed to flashes and looming stimuli responded differently than in control (F(1, 677)=25.1; p=7e-7; and F(1,629)=12.0; p=6e-4 respectively). In control neurons, longer inputs typically evoked stronger spiking, whereas neurons from stimulated animals had flatter tuning curves, with a plateau, or even a decrease in spike number for longer conductance injections (Figure 1E). In essence, while control neurons “preferred” longer inputs, stimulated neurons developed a preference for shorter synaptic inputs, and this change was more pronounced in neurons exposed to flashes than in those exposed to looming stimuli (F(1,641)=7.5; p=0.006). This suggests that the change in overall intrinsic excitability, and the change in temporal tuning, follow two different kinds of logic. The overall excitability is homeostatic, as neurons became less excitable in response to stronger stimulation. The temporal retuning however can be better described as “adaptive”, as neurons exposed to shorter stimuli (flashes) became relatively \textit{more} responsive to shorter stimuli, and less responsive to longer stimuli, which is the opposite of what one would expect for a purely homeostatic retuning. We chose to call this type of plasticity "adaptive", as presumably it means that after exposure to faster stimuli, neurons become more equipped to process faster patterns of activation \citep{stemmler1999information}.

@@

To test this question, we exposed tadpoles to four hours of behaviorally salient “click” sounds \citep{james2015,truszkowski2017}, provided at the same frequency (every second) as visual stimuli in the first set of experiments. We found (Fig 1D, E) that exposure to startle-inducing sounds (group “Sound”) did not lead to significant changes in either average spikiness (0.8$\pm$0.7 spikes; F(1,689)=1.7; p=0.2; n=28, 30), amplitude transfer function (F(1,689)=2.0; p=0.2), or temporal tuning curve (F(1,689)=2.0; p=0.2). This suggests that acoustic stimuli did not strongly activate tectal circuits during conditioning, despite being more behaviorally salient (for our stimuli, at the onset of stimulation, acoustic clicks evoked startle responses in about 50-80\% of cases, compared to 5-10\% for checkerboard inversions \citep{james2015,truszkowski2017}). This was not necessarily surprising, as auditory and mechanosensory inputs are known to have different cellular \citep{pratt2009trigeminal,felch2016} and subcellular \citep{hiramoto2009} targets in the tectum, and thus can recruit tectal networks to a different extent.

@@

To see whether intrinsic and synaptic temporal properties of tectal cells coordinated with each other, we compared intrinsic temporal tuning of every neuron (that is, whether it preferred longer or shorter simulated synaptic inputs in dynamic clamp experiments) to the actual duration of synaptic inputs it received during in-vitro stimulation of the optic chiasm. We found that, on average, cells exposed to stronger sensory stimuli preferred shorter synaptic inputs in dynamic clamp and also received shorter synaptic currents during optic chiasm stimulation, leading to a positive correlation between average synaptic and intrinsic properties for each group (Figure 3E; r=0.89, p=0.02, n=5). This means that \textit{between} treatment groups, changes in synaptic and intrinsic properties were adaptively coordinated, and the stronger the average change in synaptic transmission, the more cells reshaped their intrinsic properties to adjust to this change.

In contrast, \textit{within} experimental groups, cells that preferred shorter synaptic inputs in dynamic clamp tended to receive longer synaptic currents, and vice versa (F(1,145)=4.9, p=0.03). We calculated differences between the properties of each individual cell and the average for the experimental group to which it belonged, and showed these deviations from respective averages on one plot (Figure 3F). Cells that had shorter synaptic currents, compared to other cells in their group, tended to be selective for longer synaptic inputs in dynamic clamp (adjusted r=$-$0.19, p=0.02, n=151). This means that individual neurons tended to tune their intrinsic properties away from the typical statistics of their inputs, decreasing their responses to common input patterns, and enhancing responses to unusual patterns of synaptic activation. % \citep{stemmler1999information}.

Note that the correlation of intrinsic and synaptic properties had opposite signs between-groups (positive) and within-groups (negative), so if we lumped all cells from all groups together, we would have found no correlation between these two variables (r=$-$0.07, p=0.4, n=151). This is a textbook example of "Simpson's paradox", wherein a pattern holds within subgroups, but disappears or is reversed on a full set because of pronounced differences between groups.

For amplitude tuning, the interaction between synaptic and intrinsic parameters of tectal cells was inconclusive. The amplitude of early synaptic responses and intrinsic amplitude tuning formally correlated on a full dataset (p=0.03, r=$-$0.17, n=151), but the correlation disappeared (p$>$0.05) when the highly non-normal amplitude data was log-transformed, or when 4 extreme values (out of 135 total) were removed. When analyzed separately, the between-groups and within-groups correlations were also insignificant. A similar analysis for amplitudes of late synaptic currents also did not yield reliable results.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=3.2in]{fig4.pdf}
\caption{Electrophysiological properties of individual neurons compared to their spiking in current and dynamic clamp experiments. (\textbf{A}). A set of curves from one voltage step experiment; black bars show the areas used to average Na (bottom) and transient K (top) currents (see Methods). (\textbf{B}). Processing of ionic currents data with IV-curves translated into two parameters (threshold potential and peak current) for each ionic conductance. (\textbf{C}). Sample data from one current clamp experiment; spikes are marked with black dots. (\textbf{D}). Estimations of spikiness from current clamp experiments (horizontal axis) and dynamic clamp experiments (vertical axis) correlate. (\textbf{E}). The number of spikes registered in current clamp mode: values predicted from a linear model plotted against observed values. The model works reasonably well (61\% of variance explained). (\textbf{F}). Similar comparison for the dynamic clamp experiments: the model has very low predictive value (13\% of variance explained). Here and in E, both values are adjusted for position.}
\end{figure}

@@

In this study we show that different sensory stimuli retuned neurons in the optic tectum of \textit{Xenopus} tadpoles in different ways, thereby inducing changes in both their temporal tuning and amplitude transfer functions. This addresses our first question about the functional scope of intrinsic plasticity in the optic tectum, and shows that it goes well beyond simple adjustments of neuronal spikiness.

@@

In answer to our second question about whether intrinsic and synaptic properties of tectal cells are in any way coordinated, here we show that intrinsic and synaptic temporal properties are co-tuned in the tectum, and moreover, that this co-tuning can be modified by sensory experience. In contrast with earlier studies that report increased excitability post-stimulation \citep{aizenman2003,dong2009,ciarleglio2015}, we found that sensory stimulation led to a suppression of spiking. The reason for that, most probably, is that the visual stimulation used in earlier studies was provided with a LED box, which was so bright and of so high contrast that it caused a suppression of retinal synaptic inputs via a polyamine block of AMPA receptors \citep{aizenman2003}. This suppression then triggered a “second-order” homeostatic compensation \citep{turrigiano2011,tien2018}, making neurons spikier. In our current experiments, however, synaptic inputs were not suppressed, and neuronal activation during sensory conditioning was stronger than in control, causing a decrease in intrinsic spikiness.

>>>>

First we asked whether our stimulation protocols caused any changes in intrinsic excitability of tectal neurons. From previous studies, we knew that in tadpoles exposed to four hours of LED flashes, tectal neurons became more excitable \citep{aizenman2003,ciarleglio2015}. However, the stimuli we used in the present study were weaker, and similar to those used in behavioral experiments \citep{khakhalin2014,james2015,truszkowski2017}. We presented a checkerboard pattern that inverted once a second either instantaneously (dubbed “Flash”; Figure 1C left), or with a slow transition over the course of a second (old black squares shrank to white, while new black squares grew from old white squares, dubbed “Looming”; Figure 1C right).

@@

After conditioning, we excised the brain, performed whole cell patch clamp on tectal neurons (Figure 1A), and counted spikes produced in dynamic clamp mode in response to simulated synaptic conductances of different durations and amplitudes (Figure 1B). We used conductances of 4 different durations (100, 200, 500, and 1000 ms), and 3 different amplitudes (peak conductances of 0.2, 0.5, and 1.0 nS), matching the range of synaptic currents observed in tectal circuits \textit{in vivo} (\citealt{xu2011,khakhalin2014,ciarleglio2015}; see Methods). Based on prior research, we expected neurons exposed to flashes (checkerboard inversions) to be more excitable, but contrary to our expectations, they were on average less spiky, and generated 0.4$\pm$0.4 spikes, across all types of dynamic clamp inputs, compared to 0.9$\pm$1.0 spikes in control (Figure 1D, left column; F(1,677)=30.4, p=5e-8, n cells=28, 29). (Here and below, F-values are reported for a multivariate fixed effects analysis of variance with selected interactions, where cell id is used as a repeated measures factor; see Methods for a detailed description). Neurons exposed to looming transitions also spiked less than control neurons (0.6$\pm$0.4 spikes; F(1,629)=9.6; p=0.002; n=28, 25), but more than those exposed to instantaneous "flashes" (F(1,641)=3.7; p=4e-6; n=29, 25). This was likewise unexpected, as in the tadpole tectum, looming stimuli are known to elicit stronger responses compared to instantaneous inversions \citep{khakhalin2014}, yet the effect yielded by looming stimuli was weaker. To sum up, both the expected sign of the effect, and the relative

We then mapped the amplitude tuning of neurons (amplitude transfer function, or gain), by looking at how an increase in transmembrane conductance translated into increased spike output. Compared to control, neurons from animals exposed to visual stimuli had a flatter (constant or hill-shaped, rather than increasing) amplitude tuning curve (Fig 1D,E), as they did not increase their spiking as fast in response to larger conductances (F(1,677)=15.9, p=8e-5; and F(1,629)=8.3, p=0.004 for flash- and looming transitions respectively). The flattening of response curves seemed slightly more pronounced for flashes than for looming, but this difference was not significant (F(1,641)=3.5; p=0.06).

@@

Across input conductances of different lengths (100, 200, 500, and 1000 ms), neurons exposed to flashes and looming stimuli responded differently than in control (F(1, 677)=25.1; p=7e-7; and F(1,629)=12.0; p=6e-4 respectively). In control neurons, longer inputs typically evoked stronger spiking, whereas neurons from stimulated animals had flatter tuning curves, with a plateau, or even a decrease in spike number for longer conductance injections (Figure 1E). In essence, while control neurons preferred longer inputs, stimulated neurons developed a preference for shorter synaptic inputs, and this change was more pronounced in neurons exposed to flashes than in those exposed to looming stimuli (F(1,641)=7.5; p=0.006). This suggests that the change in overall intrinsic excitability, and the change in temporal tuning, follow two different kinds of logic. The overall excitability is homeostatic, as neurons became less excitable in response to stronger stimulation. The temporal retuning however can be better described as “adaptive”, as neurons exposed to shorter stimuli (flashes) became relatively \textit{more} responsive to shorter stimuli, and less responsive to longer stimuli, which is the opposite of what one would expect for a purely homeostatic retuning. We chose to call this type of plasticity "adaptive", as presumably it means that after exposure to faster stimuli, neurons become more equipped to process faster patterns of activation \citep{stemmler1999information}.

@@

To test this question, we exposed tadpoles to four hours of behaviorally salient “click” sounds \citep{james2015,truszkowski2017}, provided at the same frequency (every second) as visual stimuli in the first set of experiments. We found (Fig 1D, E) that exposure to startle-inducing sounds (group “Sound”) did not lead to significant changes in either average number of spikes (0.8$\pm$0.7 spikes; F(1,689)=1.7; p=0.2; n=28, 30), amplitude transfer function (F(1,689)=2.0; p=0.2), or temporal tuning curve (F(1,689)=2.0; p=0.2). This suggests that acoustic stimuli did not strongly activate tectal circuits during conditioning, despite being more behaviorally salient (for our stimuli, at the onset of stimulation, acoustic clicks evoked startle responses in about 50-80\% of cases, compared to 5-10\% for checkerboard inversions \citep{james2015,truszkowski2017}). This was not necessarily surprising, as auditory and mechanosensory inputs are known to have different cellular \citep{pratt2009trigeminal,felch2016} and subcellular \citep{hiramoto2009} targets in the tectum, and thus can recruit tectal networks to a different extent.

@@

To see whether intrinsic and synaptic temporal properties of tectal cells coordinated with each other, we compared intrinsic temporal tuning of every neuron (that is, whether it preferred longer or shorter simulated synaptic inputs in dynamic clamp experiments) to the actual duration of synaptic inputs it received, assessed by optic chiasm stimulation. We found that, \textit{on average}, in animals exposed to stronger sensory stimuli, neurons were more responsive to shorter synaptic inputs, and also received shorter synaptic currents, leading to a positive correlation between these values across experimental groups (Figure 3E; r=0.89, p=0.02, n=5). %This means that \textit{between} treatment groups, changes in synaptic and intrinsic properties were adaptively coordinated, and the stronger the average change in synaptic transmission, the more cells reshaped their intrinsic properties to adjust to this change.

In contrast, \textit{within} experimental groups, cells that preferred shorter synaptic inputs in dynamic clamp tended to receive longer actual synaptic currents, and vice versa (F(1,145)=4.9, p=0.03). The deviations of neuronal properties from their respective group averages are shown in Figure 3F. Cells that received shorter synaptic currents, compared to other cells, tended to be more responsive to longer synaptic inputs (adjusted r=$-$0.19, p=0.02, n=151). This suggests that during normal brain development, individual neurons tended to tune their intrinsic properties away from the typical statistics of their inputs, enhancing responses to unusual patterns of synaptic activation. It also means that the correlation of intrinsic and synaptic properties had opposite signs between-groups (Figure 3E; positive) and within-groups (Figure 3F; negative), which is a textbook case of a so-called "Simpson's paradox" in data analysis.

% \citep{stemmler1999information}

% , so if all cells from all groups were lumped together in one simple correlation analysis, there was correlation between these two variables (r=$-$0.07, p=0.4, n=151) (

For amplitude tuning, the interaction between synaptic and intrinsic parameters of tectal cells was inconclusive. The amplitude of early synaptic responses and intrinsic amplitude tuning formally correlated on a full dataset (p=0.03, r=$-$0.17, n=151), but the correlation disappeared (p$>$0.05) when the highly non-normal amplitude data was log-transformed, or when 4 extreme values (out of 135 total) were removed. When analyzed separately, the between-groups and within-groups correlations were also insignificant.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=3.2in]{fig4.pdf}
\caption{Electrophysiological properties of individual neurons compared to their spiking in current and dynamic clamp experiments. (\textbf{A}). A set of curves from one voltage step experiment; black bars show the areas used to average Na (bottom) and transient K (top) currents (see Methods). (\textbf{B}). Processing of ionic currents data with IV-curves translated into two parameters (threshold potential and peak current) for each ionic conductance. (\textbf{C}). Sample data from one current clamp experiment; spikes are marked with black dots. (\textbf{D}). Estimations of excitability from current clamp experiments (horizontal axis) and dynamic clamp experiments (vertical axis) correlate. (\textbf{E}). The number of spikes registered in current clamp mode: values predicted from a linear model plotted against observed values. The model works reasonably well (61\% of variance explained). (\textbf{F}). Similar comparison for the dynamic clamp experiments: the model has very low predictive value (13\% of variance explained). Here and in E, both values are adjusted for position.}
\end{figure}

@@

In this study we show that different sensory stimuli retuned neurons in the optic tectum of \textit{Xenopus} tadpoles in different ways, thereby inducing changes in both their temporal tuning and amplitude transfer functions. This addresses our first question about the functional scope of intrinsic plasticity in the optic tectum, and shows that it goes well beyond simple adjustments of neuronal excitability.

@@

In answer to our second question about whether intrinsic and synaptic properties of tectal cells are in any way coordinated, here we show that intrinsic and synaptic temporal properties are co-tuned in the tectum, and moreover, that this co-tuning can be modified by sensory experience. In contrast with earlier studies that report increased excitability post-stimulation \citep{aizenman2003,dong2009,ciarleglio2015}, we found that sensory stimulation led to a suppression of spiking. The reason for that, most probably, is that the visual stimulation used in earlier studies was provided with a LED box, which was so bright and of so high contrast that it caused a suppression of retinal synaptic inputs via a polyamine block of AMPA receptors \citep{aizenman2003}. This suppression then triggered a “second-order” homeostatic compensation \citep{turrigiano2011,tien2018}, making neurons spikier. In our current experiments, however, synaptic inputs were not suppressed, and neuronal activation during sensory conditioning was stronger than in control, causing a decrease in intrinsic excitability.
